Hi o3

At the next prompt, I am going to provide you with entries from a spreadsheet describing digital resources for archaeology.

We need to improve the data in the following ways:

1. improve its description to a 50-100 word description
2. decide if the following tool definition applies 

<definition>

For our project, a software tool is any discrete piece of software—be it a program, script, or web application—that is developed or substantially modified by researchers to perform specific, research-oriented computational tasks. These tasks must involve active data processing, such as analysis, simulation, visualization, modeling, or automation, where input data is dynamically created, documented, transformed, or interpreted to generate meaningful results. In contrast, products whose primary function is to simply display, store, or promote static datasets—without offering mechanisms for active computational engagement—are excluded. 

In essence, a valid software tool for our purposes is one that not only **supports research in archaeology or historical sciences** but does so by actively collecting and/or processing data rather than merely serving as a generic data repository or static interface.

</definition>

If the resource does meet this definition, please: 

1. Add a new column named 'archhist-software-tool' to the spreadsheet if it does not already exist.
2. Place a 'yes' or a 'no' in that column to indicate whether or not the resource meets our definition of a tool. 

Output:

Please output the following columns: 'item_name' (unchanged from original), 'description' (with the original description), 'Improved description' (with your improved description), and 'archist-software-tool' as a CSV in a fenced code block. 

Do you understand these instructions?